---
title: "Pilotstudy"
author: ""
date: ""
output:
  pdf_document: default
  html_document: default

---

## Implementation

- [Orignal paper](https://osf.io/8vcxh/overview)

#### Setup \

Before running the script of your choice (either a .R or a .Rmd script) check that you defined the data generating function *dgp(beta1, restrict, n)*, the parameter values for *beta1* and *n*, and the model formula and received the Helper.R script. This script incorporates multiple validation checks for the confirmatory study to verify that all restrictions are met and that every component is correctly specified. It additionally provides a fixed seed for the simulations and contains the complete code used to perform the simulations and compute the effect size estimates, relative bias and power. Also code to represent the results is provided in this script. \

You are encouraged to include comments in your code, especially in the exploratory section. \
If the script does not run and there are no error messages check again the upper rules for the \texttt{dgp(beta1, restrict, n)} function. \
If you do not find the problem, feel free to contact us: [multi.analyst\@stat.uni-muenchen.de](mailto:multi.analyst@stat.uni-muenchen.de){.email}\


```{r setup}
# load helper functions into the current R session
# make sure the file is in you current working direction
source("Helper.R")
```


#### DGP 

```{r dgp}
dgp <- function(beta1, restrict = c("none", "weak", "moderate", "strong"), n, ...){
  # Template: implement the data-generator and return two data.frames with columns:
  #   - x : numeric or binary predictor
  #   - y : numeric outcome
  # Additional variables (covariates) may be returned as extra columns.
  #
  # define the DGP
  

  # no restriction
    # if(restrict == "none"){
    #  convenience_x <- convenience_y <- ...
    # }

  # restrict by x 
  
    # define weak restriction: 
    # if(restrict == "weak"){
    # ...
    # }
  
    # define moderate restriction: 
    # if(restrict == "moderate"){
    # ...
    # }
  
    # define strong restriction: 
    # if(restrict == "strong"){
    # ...
    # }
  
  # restrict by y
  
    # define weak restriction: 
    # if(restrict == "weak"){
    # ...
    # }
  
    # define moderate restriction: 
    # if(restrict == "moderate"){
    # ...
    # }
  
    # define strong restriction: 
    # if(restrict == "strong"){
    # ...
    # }

  

  # list(convenience_x = , convenience_y =) 
  # keep this and include your data.frame behind the =
}
```


#### Effect size

```{r eff_size}
beta1 <- c(NA, NA, NA)
```

#### Number of observations per sample

```{r n}
n <- NA
```


#### Model formula

```{r model_formula}
# formula <- y ~ x + cov1 + cov2 
# your data.frames from dgp should include these variables!
```


## Results of the Monte Carlo design \

This section repeats your data-generating procedure for $n_{iter}$ Monte Carlo iterations (defined as *niter*; by default this is set to 10,000, although we recommend using 100 for testing purposes).
This will estimate the variability and bias of your effect size measures across repeated samples. To obtain the results, simply run the sections below. \
**Run the checks:** Before running the full simulation, execute the *checks()* function.  
If no errors or warnings occur, your DGP is valid and you can proceed with the simulation.

**Run the simulation:** Once the checks pass, use the *run\_simulations()* function to perform the Monte Carlo iterations. The wrapper collects per-iteration results and then computes summary quantities across iterations (mean relative bias, Monte Carlo standard error of the mean, and 95\% simulation interval).\

```{r iterate}
# suggested number for reliable results niter = 10000
niter <- 100 # number of iterations for test purposes

# run checks before running the simulation
# if no error occurs you can run the simulation sim
checks(niter, beta1, restrict, n, formula, dgp)

# run if you got the go from the checks()-function
sim <- run_simulations(niter, beta1, restrict, n, formula, dgp)
```


**Represent the results:** Run the two following functions. *simulation_results()* returns two
figures showing the results for sample restriction of *X* and *Y*, respectively. \
Note: the Pearson correlation coefficient is only derived for a numeric *X* and the cohen’s d only for a binary *X*.
The function *power()* returns a figure displaying the power for the unrestricted sample and the convenience samples.
All figures will be saved in the same folder where your Helper.R script is located. Please send us your completed script along with the three resulting figures and the *sessionInfo()* text file generated at the end of your script. \

```{r render_summary}
simulation_results(sim, restrict)
# loss in power
power(sim)
```


## Exploratory part
In the exploratory part, you are free to extend the confirmatory simulations with any settings you
consider relevant or interesting. There are no restrictions on the data-generating process (DGP) or
the types of violations you introduce (e.g., model misspecification, nonlinearity, measurement error, etc.). Please document your implementation and design choices in a short written report. \
You may, however, reuse the functions in Helper.R to the extent that your exploratory design remains compatible with their assumptions. \

**When you can use run_simulations():** The function run_simulations() can be used as long
as the following arguments are specified in the required format: \
- niter: an integer giving the number of Monte Carlo iterations. \
- beta1: a numeric vector of length three, specifying the true effect sizes to be evaluated. \
- restrict: a character vector containing the four restriction levels "none", "weak", "moderate", and
"strong". \
- n: an integer specifying the sample size per simulated dataset. \
- formula: an R formula specifying the analysis model (e.g., y x + z). \
- dgp_fun: a data-generating function of the form dgp_fun(beta1, restrict, n) as described above,
which returns the required samples (unrestricted, restricted on X, restricted on Y) in the same
structure as in the confirmatory part. \
Within this framework, you may modify the DGP quite freely, for example, by introducing het-
eroscedastic errors, nonlinear terms, effect modification, or measurement error, provided that the
arguments and output structure expected by run_simulations() and one_simulation_run() are pre-
served. \

**When the Helper functions are no longer sufficient:** The current Helper functions are tai-
lored to the effect size measures used in the confirmatory part (unstandardized and standardized
regression coefficients, Pearson’s r, Cohen’s d, R2) and the corresponding confidence intervals and
power. If, in your exploratory analyses, you wish to: \
- derive additional effect size measures (e.g., Spearman rank correlation, odds ratios, partial cor-
relations), or \
- focus on other performance metrics (e.g., predictive performance, calibration measures, classifi-
cation accuracy), \
then you will need to implement the computation of these quantities yourself (e.g., by extending
model_function(), adding new summary code, or writing separate analysis functions). In such cases,
run_simulations() and the plotting functions (simulation_results(), power()) may serve as templates,
but will not automatically handle the new measures without modification.

```{r exploratory}

```


## Reproducibility information (required): \

Please run the following code at the end of your script. 
This will generate a .txt file containing your R version and package versions. 
Please submit this file together with your script and outputs.
```{r sessionInfo}
writeLines(
  capture.output(sessionInfo()),
  paste0("sessionInfo_", Sys.Date(), ".txt")
)
```
